Species-specific effects of the Urban Heat Island on tree growth across Berlin

### 2020-03-20 analysis/paper/paper.Rmd
cont. writing

---
title: "Species-specific effects of the Urban Heat Island on tree growth across Berlin"
author:
  - Alexander G. Hurley$^{a *}$, Ingo Heinrich$^a$
  - $^a$ *Geoforschungszentrum Potsdam, Section 4.3, Germany*
  - "* Corresponding author: hurley@gfz-potsdam.de"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::html_document2:
      fig_caption: yes
      toc: true
    bookdown::pdf_document2:
      fig_caption: yes
      toc: yes
bibliography: uhi_references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  **This document serves as a brief overview and outline** .
keywords: |
  Uferbaeume; Anlagenbaeume; Strassenbaeume; 
highlights: |
  We will see. 
---

@@

Conditions affecting tree growth can vary greatly within urban areas or regions, and need to be accounted for when establishing relationships with pertinent drivers, such as the UHI effect.
This typically complicates the extrapolation from individual sampling sites toward predicting effect sizes across entire urban areas and tree stocks.
This is especially the case for studies reliant on labour-intensive methods which are limited logistically by sampling effort, reducing sample sizes, as well as species and spatial coverage.
<!-- While multiple (locally) important species have been assessed along urban and UHI gradients reflecting a range of conditions, studies reliant on labour-intensive methods are limited logistically by sampling effort, reducing sample size and coverage (i.e. genera and/or species, and space). -->
<!-- However, urban tree growth in urban settings is affected by various processes, which  -->
<!-- However, growth in urban settings is affected by other processes and factors, such as water availability, pollution and road-salt loading, structural impedance through infrastructure or management, etc. [@rhoades1999; @quigley2004; @randrup2001; @pauleit2002]. -->
<!-- While multiple (locally) important species have been assessed along urban and UHI gradients reflecting a range of conditions, studies reliant on labour-intensive methods are limited logistically by sampling effort, reducing sample size and coverage (i.e. genera and/or species, and space). -->
<!-- This typically hinders the extrapolation from individual sampling sites toward prediction of effect sizes across entire urban areas and tree stocks.  -->
<!-- Therefore, studies reliant on labor-intensive sampling, must carefully address confounding factors to ensure effect sizes are estimated adequately. -->
To complement detailed dendroecological analyses of climate-growth relationships in Berlin for key species, we propose inferring growth modulation from a large data set in excess of 650000 individuals provided by the Berlin Senate Administration (Senatsverwaltung). 
This data set contains information on location, species, trunk diameter (at breast height; $DBH$), and height, amongst other variables for street and park trees.
In a space-for-time substitution, growth of individual species can be assessed across the entire cite of Berlin, and related to effects of the UHI, while accounting for other location-specific factors, such as street characteristics, development intensity, available soil volume, etc.
Comparable applications are found, for example, in @quigley2004 and @pretzsch2017.
The former  inferred absolute growth potential for species across successional groups, and between rural and urban species, yet lacked spatially-explicit effect size estimates or predictions of maximum potential;
@pretzsch2017 applied linear hierarchical models to infer growth modulation for different cities, time periods and locations (urban $vs.$ rural).

**By contrast, we propose applying a statistical model that is spatially explicit, while also allowing to account for the nested nature of the data set (e.g. streets and districts) as well as other pertinent factors (hierarchical, generalized additive model, see Section$~$\@ref(sec:methods)).
This also allows to infer the absolute growth potential of species given, for example, a specific location, age or UHI magnitude.**

@@

\begin{equation}
E (Y)~=~g^{-1}\left( \beta_0 + \sum_{i = 1}^{n} f_i (x_i) \right),
\end{equation}

@@

\begin{equation}
y~=~E (Y) + \epsilon,
\end{equation}

@@

\begin{equation}
f_i(x_i)~=~\sum_{k = 1}^{K} \beta_{i, k} b_{i,k}(x_i).
\end{equation}

@@

\begin{equation}
Y_{i,j} = (\beta_0 + b_{0,i,j}) + (\beta_1 + b_{1,i,j}) \cdot x_i  + \epsilon_{i,j},
\end{equation}
where $\beta_0$ is the intercept with its random component $b_0$, and $\beta_1$ the slope with its random component $b_1$.
The random errors are assumed i.i.d. and distributed as $b \sim \mathcal{N}(0, \tau ^2)$. 
The model for which results are presented in Figure$~$\@ref(fig:fig-lmestat) estimates $DBH$ from tree age and the local UHI intensity as continuous covariates with random slopes and intercepts for each species;
note, that for computational efficiency each genera was modelled seperately.
Further, models were only established for species with at least 150 individuals.

@@

             caption = "Data requirements for analysis, including currently available/accessed and required/desired data.",
             caption.short = "Data requirements for analysis",
             col.names = c("Subject/Relevance",
                          "Desc.",
                          "Type",
                          "Obsv. (n)",
                          "Resolution (m)",
                          "Source")) %>% 
  kableExtra::kable_styling(full_width = TRUE, 
                            font_size = 10) %>% 
  kableExtra::row_spec(0, bold = TRUE) %>% 
  kableExtra::pack_rows("available or accessed", 1, 4) %>%
  kableExtra::pack_rows("required and/or desired", 5, 8) %>% kableExtra::landscape()

} else {
  table_dat_req <- dat_req %>% 
  select(-owned) %>% 
  knitr::kable(booktabs = TRUE,
               format = "html",
               caption = "Data requirements for analysis, including currently available/accessed and required/desired data.",
             caption.short = "Data requirements for analysis",
             col.names = c("Subject/Relevance",
                          "Desc.",
                          "Type",
                          "Obsv. (n)",
                          "Resolution (m)",
                          "Source")) %>% 
  kableExtra::kable_styling(full_width = TRUE, 
                            font_size = 10) %>% 
  kableExtra::row_spec(0, bold = TRUE) %>% 
  kableExtra::pack_rows("available or accessed", 1, 4) %>%
  kableExtra::pack_rows("required and/or desired", 5, 8)

@@

Some results

@@

@@

```{r fig-density, fig.cap="Empirical density distribution of all individuals within the presented genera along the UHI continuum. UHI intensities were extracted for each tree location, and the distribution hence represents the first detailed overview of the exposure of Berlin's trees to urban heat loading. Insets represent corresponding tree totals.",fig.width = 12, fig.height = 12, out.width="100%"}
knitr::include_graphics(here::here("analysis/figures/plot_02_genus_UHI_dens.png"))

@@

@@

- incorporating more pertinent covariates as dependent variables in the linear mixed model
- testing multiple model structures with formal model selection procedures
- checking model resiudals for spatial auto-correlation and accounting for it where necessary to ensure unbiased estimates of effect sizes
- repeating the above with a hierarchical GAM (i.e. GAMM) to allow for:  
  * estimating continuous prediction surfaces for UHI impacts on individual species' growth (similar to results in Figure$~$\@ref(fig:fig-lmestat)) under recent conditions
  * estimating absolute, species-specific growth potential under increased temperatures and UHI loading under climate change, ideally based on simulations (otherwise step-wise increases based on RCP scenarios) for the key species.
  * assess potential age-dependent UHI impacts on individual species.

>>>>

---
title: "Species-specific effects of the Urban Heat Island on tree growth across Berlin"
author:
  - Alexander G. Hurley$^{a *}$, Ingo Heinrich$^a$
  - $^a$ *Geoforschungszentrum Potsdam, Section 4.3, Germany*
  - "* Corresponding author: hurley@gfz-potsdam.de"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::html_document2:
      fig_caption: yes
      toc: true
    bookdown::pdf_document2:
      fig_caption: yes
      toc: yes
bibliography: uhi_references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  **This document outlines the rational for an analysis tree growth (potential) and its relationship with the Urban Heat Island (UHI) effect in Berlin. It introduces preliminary results and provides an outlook for up-coming and potential work.** .
keywords: |
  Uferbaeume; Anlagenbaeume; Strassenbaeume; 
highlights: |
  We will see. 
---

@@

Conditions affecting tree growth can vary greatly within urban areas or regions, and need to be accounted for when establishing relationships with pertinent drivers, such as the UHI effect.
This typically complicates the extrapolation from individual sampling sites toward predicting effect sizes across entire urban areas and tree stocks.
This is especially the case for studies reliant on labour-intensive methods which are limited logistically by sampling effort, reducing sample sizes, as well as species and spatial coverage.
<!-- While multiple (locally) important species have been assessed along urban and UHI gradients reflecting a range of conditions, studies reliant on labour-intensive methods are limited logistically by sampling effort, reducing sample size and coverage (i.e. genera and/or species, and space). -->
<!-- However, urban tree growth in urban settings is affected by various processes, which  -->
<!-- However, growth in urban settings is affected by other processes and factors, such as water availability, pollution and road-salt loading, structural impedance through infrastructure or management, etc. [@rhoades1999; @quigley2004; @randrup2001; @pauleit2002]. -->
<!-- While multiple (locally) important species have been assessed along urban and UHI gradients reflecting a range of conditions, studies reliant on labour-intensive methods are limited logistically by sampling effort, reducing sample size and coverage (i.e. genera and/or species, and space). -->
<!-- This typically hinders the extrapolation from individual sampling sites toward prediction of effect sizes across entire urban areas and tree stocks.  -->
<!-- Therefore, studies reliant on labor-intensive sampling, must carefully address confounding factors to ensure effect sizes are estimated adequately. -->

<!-- Detailed, but spatially-limited dendroecological analyses of climate-growth relationships in Berlin for key species can be enhanced by inferring growth modulation from a large data set in excess of 650000 individuals provided by the Berlin Senate Administration (Senatsverwaltung). -->

To complement (existing) detailed dendroecological analyses of climate-growth relationships in Berlin for key species, we propose inferring growth modulation from a large data set in excess of 650000 individuals provided by the Berlin Senate Administration (Senatsverwaltung).
This data set contains information on location, species, trunk diameter (at breast height; $DBH$), and height, amongst other variables for street and park trees.
In a space-for-time substitution, growth of individual species can be assessed across the entire cite of Berlin, and related to effects of the UHI, while accounting for other location-specific factors, such as street characteristics, development intensity, available soil volume, etc.
Comparable applications are found, for example, in @quigley2004 and @pretzsch2017.
The former  inferred absolute growth potential for species across successional groups, and between rural and urban species, yet lacked spatially-explicit effect size estimates or predictions of maximum potential;
@pretzsch2017 applied linear hierarchical models to infer growth modulation for different cities, time periods and locations (urban $vs.$ rural).

**By contrast, we propose applying a statistical model that is spatially explicit, while also allowing to account for the nested nature of the data set (e.g. streets and districts) as well as other pertinent factors (hierarchical, generalized additive model, see Section$~$\@ref(sec:methods)).
This also allows to infer the absolute growth potential of species given, for example, a specific location, age or UHI magnitude.**
Tree-level growth data, however, is paramount in validating such relationships, and its inclusion in the model would also allow incorporating effects of varying climate.

@@

\begin{equation}
E (Y)~=~g^{-1}\left( \beta_0 + \sum_{i = 1}^{n} f_i (x_i) \right),
(\#eq:gam-general)
\end{equation}

@@

\begin{equation}
y~=~E (Y) + \epsilon,
(\#eq:gam-expectation)
\end{equation}

@@

\begin{equation}
f_i(x_i)~=~\sum_{k = 1}^{K} \beta_{i, k} b_{i,k}(x_i).
(\#eq:gam-smoothers)
\end{equation}

@@

\begin{equation}
Y_{i,j} = (\beta_0 + b_{0,i,j}) + (\beta_1 + b_{1,i,j}) \cdot x_i  + \epsilon_{i,j},
(\#eq:lme)
\end{equation}
where $\beta_0$ is the intercept with its random component $b_0$, and $\beta_1$ the slope with its random component $b_1$.
The random errors are assumed i.i.d. and distributed as $b \sim \mathcal{N}(0, \tau ^2)$. 
The model for which results are presented in Figure$~$\@ref(fig:fig-lmestat) estimates $DBH$ from tree age and the local UHI intensity as continuous covariates with random slopes and intercepts for each species;
note, that for computational efficiency each genera was modelled seperately.
Further, models were only established for species with at least 150 individuals.

@@

             caption = "Data requirements for analysis, including currently available/accessed and required/desired data.",
             caption.short = "Data requirements for analysis",
             col.names = c("Subject/Relevance",
                          "Desc.",
                          "Type",
                          "Obsv. (n)",
                          "Resolution (m)",
                          "Source")) %>% 
  kableExtra::kable_styling(full_width = TRUE, 
                            font_size = 11) %>% 
  kableExtra::row_spec(0, bold = TRUE, font_size = 12) %>% 
  kableExtra::row_spec(9,italic =  TRUE) %>% 
  kableExtra::pack_rows("available or accessed", 1, 4) %>%
  kableExtra::pack_rows("required and/or desired", 5, 9) %>% kableExtra::landscape()

} else {
  table_dat_req <- dat_req %>% 
  select(-owned) %>% 
  knitr::kable(booktabs = TRUE,
               format = "html",
               caption = "Data requirements for analysis, including currently available/accessed and required/desired data.",
             caption.short = "Data requirements for analysis",
             col.names = c("Subject/Relevance",
                          "Desc.",
                          "Type",
                          "Obsv. (n)",
                          "Resolution (m)",
                          "Source")) %>% 
  kableExtra::kable_styling(full_width = TRUE, 
                            font_size = 10) %>% 
  kableExtra::row_spec(0, bold = TRUE) %>% 
  kableExtra::pack_rows("available or accessed", 1, 4) %>%
  kableExtra::pack_rows("required and/or desired", 5, 9)

@@

Tree locations are clustered and follow spatial structures based on their category, i.e. riparian, street and park trees (Fig.$~$\@ref(fig:fig-tree-overview-map)).
Planting in space and time shows species specific patterns (by districts), often related to major events, such as the start and end of wars, as well as the German separation (not shown here).
Table$~$\@ref(tab:tab-age) shows the binned distribution of generae across age classes.

```{r tab-age}

loadd(age_tables)

age_tables <- age_tables %>% 
  dplyr::arrange(desc(n_total))
colnames(age_tables)[c(1,7,8)] <- c("Genera", "Total (n)", "Missing (n)")

age_tables <- bind_rows(age_tables[age_tables$Genera!="Other",],
                        age_tables[age_tables$Genera=="Other",])

add_row <- as.data.frame(matrix(c(NA, colSums(age_tables[,-1])), nrow = 1))
add_row[1,1] <- "Marg. Totals"
colnames(add_row) <- colnames(age_tables)

age_tables <- bind_rows(age_tables,
                    add_row)

if (grepl("docx", out_format) ||
    knitr::is_latex_output() ) {
  table_dat_req_kable <- age_tables %>% 
  knitr::kable(booktabs = TRUE,

             caption = "Trees",
             caption.short = "Trees"
             # col.names = c("Subject/Relevance",
             #              "Desc.",
             #              "Type",
             #              "Obsv. (n)",
             #              "Resolution (m)",
             #              "Source")
             ) %>% 
  kableExtra::row_spec(13, hline_after = TRUE)

} else {
table_dat_req_kable <- age_tables %>% 
  knitr::kable(booktabs = TRUE,
               format = "html",
               caption = "Trees",
             caption.short = "Trees"
             # col.names = c("Subject/Relevance",
             #              "Desc.",
             #              "Type",
             #              "Obsv. (n)",
             #              "Resolution (m)",
             #              "Source")
             ) %>% 
  kableExtra::row_spec(13, extra_css = "border-bottom: 1px solid")

# %>% 
#   kableExtra::kable_styling(full_width = TRUE, 
#                             font_size = 10) %>% 
#   kableExtra::column_spec(1, italic = TRUE) %>%
#   kableExtra::row_spec(0, bold = TRUE) %>% 
#     kableExtra::row_spec(13, hline_after = TRUE)

   table_dat_req_kable %>% 
  kableExtra::kable_styling(full_width = TRUE, 
                            font_size = 11) %>% 
  kableExtra::column_spec(1, italic = TRUE) %>%  
  kableExtra::row_spec(0, italic = FALSE, bold = TRUE, font_size = 12)

@@

The distribution of the UHI effect is highly irregular and clustered in space (Fig.$~$\@ref(fig:fig-uhi-map)), and also shows variability through time (data not shown, refer to the [urban heat island explorer](https://yceo.users.earthengine.app/view/uhimap)).

@@

The exposure to increased heat-loading of individual genera (and consequently species) is highly uneven throughout the city (Fig.$~$\@ref(fig:fig-density)).
Street and park trees of most genera are clustered in urban areas with intermediate to high UHI loading, while riparian trees, and some street and park trees of other genera tend to be spread more evenly across Berlin's UHI range.

```{r fig-density, fig.cap="Empirical density distribution of all individuals within the presented genera along the UHI continuum. UHI intensities were extracted for each tree location, and the distribution hence represents the first detailed overview of the exposure of Berlin's trees to urban heat loading. The black line represents the density across all three categories. Insets represent corresponding tree totals.",fig.width = 12, fig.height = 12, out.width="100%"}
knitr::include_graphics(here::here("analysis/figures/plot_02_genus_UHI_dens.png"))

@@

The effect of UHI loading on absolute growth potential varies between genera and species (Fig.$~$\@ref(fig:fig-lmestat)).
Most notably, *Quercus*, the 3rd-most frequent genera, shows decreased absolute growth with increased UHI,
while the most frequent genera, *Tilia* clearly features enhanced growth.
Note, that the estimated effect sizes here are linear.
Temperature may excert a non-linear control on absolute growth, however, and hence, applying a  method able to capture may result in altered presented effect sizes.
Additionally, if temperatures increase in the future under climate warming current future increases in under climate warming, any non-linear effects may become more enhanced, stressing the need for a more flexible model fit and structure.

alter the (preliminary) results substantially

@@

- incorporating more pertinent covariates as dependent variables in the linear mixed model
- testing multiple model structures with formal model selection procedures
- checking model resiudals for spatial auto-correlation and accounting for it where necessary to ensure unbiased estimates of effect sizes
- repeating the above with a hierarchical GAM (i.e. GAMM) to allow for:  
  * estimating continuous prediction surfaces for UHI impacts on individual species' growth (similar to results in Figure$~$\@ref(fig:fig-lmestat)) under recent conditions
  * estimating absolute, species-specific growth potential under increased temperatures and UHI loading under climate change, ideally based on climate simulations (otherwise step-wise increases based on RCP scenarios) for the key species. Note, that  complications presented by 'out-of-sample' predictions will be addressed.
  * assess potential age-dependent UHI impacts on individual species.
